---
title: "RW_Group8"
output: html_document
date: "2026-02-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 5)
library(arm)
library(car)
library(ggplot2)
df <- read.csv("RWNS_final.csv", stringsAsFactors = FALSE)
df <- subset(df, !is.na(ks4score))
```

#Explore the data using plots to analyse which categorical variabes can be merged 
```{r}
ggplot(df, aes(x = gender, y = ks4score, fill = gender)) +
  geom_boxplot() +
  labs(title = "KS4 Score Across Different Genders", x = "Gender", y = "KS4 Score") +
  theme_minimal()

# mother's education
ggplot(df, aes(x = hiquamum, y = ks4score, fill = hiquamum)) +
  geom_boxplot() +
  scale_x_discrete(guide = guide_axis(n.dodge = 2)) +
  labs(title = "KS4 Score Based On Mother's Qualification Status", x = "Mother's Qualification", y = "KS4 Score") +
  theme_minimal()

# social economies class
ggplot(df, aes(x = SECshort, y = ks4score, fill = SECshort)) +
  geom_boxplot() +
  labs(title = "KS4 Scores Across Different Social Classes", x = "SEC Category", y = "KS4 Score") +
  theme_minimal()

# computers
ggplot(df, aes(x = computer, y = ks4score, fill = computer)) +
  geom_boxplot() +
  labs(title = " KS4 Score Across Home Computer Ownership", x = "Has Computer", y = "KS4 Score") +
  theme_minimal()

# school meals
ggplot(df, aes(x = fsm, y = ks4score, fill = fsm)) +
  geom_boxplot() +
  labs(title = "KS4 Score Across Free School Meal Status", x = "FSM Entitled", y = "KS4 Score") +
  theme_minimal()

# attitude
ggplot(df, aes(x = attitude, y = ks4score, fill = fsm)) +
  geom_boxplot() +
  labs(title = "KS4 Score Across different attitudes", x = "Attitudes", y = "KS4 Score") +
  theme_minimal()

# homework
ggplot(df, aes(x = homework, y = ks4score, fill = fsm)) +
  geom_boxplot() +
  labs(title = "KS4 Score Across different homework status", x = "Homework", y = "KS4 Score") +
  theme_minimal()

```

#Convert missing strings to explicit level
```{r}
for (v in c("gender","singlepar","house","fsm","computer","pupasp",
            "parasp","truancy","absent","exclude","sen","tuition","SECshort","hiquamum")) {
  x <- as.character(df[[v]])
  x[x %in% c("-1","-9","missing")] <- "Missing"
  df[[v]] <- factor(x)
}
```


#Combine the variables we analyzed and take highly correlated variables into consideration
```{r eda_cor}
## Confirms high correlation between KS3 subscores -> justifies using ks3_mean
cor(df[, c("k3en","k3ma","k3sc")], use = "complete.obs")
```

```{r load}

## Combine KS3 subscores into one mean (k3en/k3ma/k3sc are highly correlated)
df$ks3_mean <- rowMeans(df[, c("k3en", "k3ma", "k3sc")], na.rm = TRUE)
#Combine mother's qualification
df$mum_edu3 <- as.character(df$hiquamum) 
df$mum_edu3[grepl("Degree|HE", df$hiquamum, ignore.case = TRUE)] <- "Degree/HE"
df$mum_edu3[grepl("None|Other|missing", df$hiquamum, ignore.case = TRUE)] <- "None/Other/Missing"
df$mum_edu3 <- factor(df$mum_edu3) 

#social status combine
df$sec3 <- as.character(df$SECshort)
df$sec3[grepl("Routine|missing", df$SECshort, ignore.case = TRUE)] <- "Lower/Missing"
df$sec3[grepl("Managerial", df$SECshort, ignore.case = TRUE)] <- "Professional"
df$sec3[grepl("Intermediate", df$SECshort, ignore.case = TRUE)] <- "Intermediate"
df$sec3 <- factor(df$sec3)
  
## attitude: merge 4 levels into 2
df$attitude3 <- ifelse(df$attitude %in% c("very_high", "high"), "high",
                ifelse(df$attitude %in% c("low", "very_low"), "low", "Missing"))
df$attitude3 <- factor(df$attitude3, levels = c("high", "low", "Missing"))

## homework: merge 6 levels into 3
df$homework3 <- ifelse(df$homework %in% c("none", "1_evening"),       "0-1",
                ifelse(df$homework %in% c("2_evenings", "3_evenings"), "2-3",
                ifelse(df$homework %in% c("4_evenings", "5_evenings"), "4-5",
                "Missing")))
df$homework3 <- factor(df$homework3, levels = c("0-1", "2-3", "4-5", "Missing"))


## Set reference levels
df$gender    <- relevel(df$gender,    ref = "Female")
df$singlepar <- relevel(df$singlepar, ref = "no")
df$house     <- relevel(df$house,     ref = "owned")
df$fsm <- as.character(df$fsm)
df$fsm[df$fsm == "Missing"] <- "no"
df$fsm <- relevel(factor(df$fsm), ref = "no")
df$truancy   <- relevel(df$truancy,   ref = "No")
df$absent    <- relevel(df$absent,    ref = "No")
df$exclude   <- relevel(df$exclude,   ref = "No")
df$sen       <- relevel(df$sen,       ref = "No")
df$pupasp    <- relevel(df$pupasp,    ref = "Yes")
df$parasp    <- relevel(df$parasp,    ref = "Yes")
df$sec3      <- relevel(df$sec3,      ref = "Professional")
df$mum_edu3  <- relevel(df$mum_edu3,  ref = "Degree/HE")
```

# 2. Exploratory Data Analysis
```{r eda_scatter}
with(df, plot(ks3_mean, ks4score,
              pch = 16, cex = 0.3, col = rgb(0,0,0,0.2),
              xlab = "Mean KS3 level", ylab = "KS4 score",
              main = "KS4 vs KS3 mean"))
abline(lm(ks4score ~ ks3_mean, data = df), col = "red")
```


#Boxplot after merging and remove missing values
```{r eda_boxplots, fig.height=9}
par(mfrow = c(3, 3))
boxplot(ks4score ~ gender,    data = df, main = "Gender",             ylab = "KS4 score")
boxplot(ks4score ~ SECshort,  data = df, main = "Parental SEC",       ylab = "KS4 score", las=2, cex.axis=0.6)
boxplot(ks4score ~ mum_edu3,  data = df, main = "Mother's qual",      ylab = "KS4 score", las=2, cex.axis=0.6)
boxplot(ks4score ~ fsm,       data = df, main = "Free school meal",   ylab = "KS4 score")
boxplot(ks4score ~ attitude3, data = df, main = "Attitude to school", ylab = "KS4 score")
boxplot(ks4score ~ homework3, data = df, main = "Homework frequency", ylab = "KS4 score")
boxplot(ks4score ~ truancy,   data = df, main = "Truancy",            ylab = "KS4 score")
boxplot(ks4score ~ sen,       data = df, main = "SEN",                ylab = "KS4 score")
boxplot(ks4score ~ pupasp,    data = df, main = "Pupil aspiration",   ylab = "KS4 score")
boxplot(ks4score ~ sec3,    data = df, main = "Economic and social status",   ylab = "KS4 score")
par(mfrow = c(1, 1))
```

# 3. Regression Analysis
## Full model
```{r full_model}
m_full <- lm(ks4score ~ ks3_mean + gender + singlepar + house + IDACI_n +
               computer + pupasp + parasp + attitude3 + homework3 +
               truancy + absent + exclude + sen +
               fsm + SECshort + hiquamum + tuition,
             data = df)
display(m_full)
summary(m_full)
```

## Backward elimination
```{r backward}
## Remove non-significant predictors one group at a time (p > 0.05)
## Step 1: remove 'computer' - all levels non-significant
m1 <- update(m_full, . ~ . - computer)

## Step 2: remove 'tuition' - non-significant
m2 <- update(m1, . ~ . - tuition)

## Step 3: remove 'absent' - non-significant once other factors controlled
m3 <- update(m2, . ~ . - absent)

## Step 4: remove 'singlepar' - non-significant once SEC and house included
m4 <- update(m3, . ~ . - singlepar)

## Step 5: remove 'SECshort' - all levels non-significant (p > 0.3)
m_final <- update(m4, . ~ . - SECshort)

display(m_final)
summary(m_final)
```

#check the r^2 value
```{r}
get_diag <- function(model, name) {
  cat("\n--- Diagnostics for", name, "---\n")
  cat("Adj R-squared:", summary(model)$adj.r.squared, "\n")
  cat("AIC:", AIC(model), "\n")
}

get_diag(m_full, "Full Model")

## Step 1: remove 'computer'
m1 <- update(m_full, . ~ . - computer)
get_diag(m1, "m1 (removed computer)")

## Step 2: remove 'tuition' 
m2 <- update(m1, . ~ . - tuition)
get_diag(m2, "m2 (removed tuition)")

## Step 3:remove 'absent'
m3 <- update(m2, . ~ . - absent)
get_diag(m3, "m3 (removed absent)")

## Step 4:remove 'singlepar' 
m4 <- update(m3, . ~ . - singlepar)
get_diag(m4, "m4 (removed singlepar)")

## Step 5: remove 'SECshort'
m_final <- update(m4, . ~ . - SECshort)
get_diag(m_final, "Final Model")


summary(m_final)

par(mfrow = c(2, 2))
plot(m_final)
par(mfrow = c(1, 1))
```


# 4. Diagnostics

```{r diagnostics, fig.height=6}
par(mfrow = c(2, 2))
plot(m_final)
par(mfrow = c(1, 1))
```

```{r outliers}
show_outliers <- function(the.linear.model, topN) {
  n <- length(fitted(the.linear.model))
  p <- length(coef(the.linear.model))
  res.out    <- which(abs(rstandard(the.linear.model)) > 3)
  res.top    <- head(rev(sort(abs(rstandard(the.linear.model)))), topN)
  lev.out    <- which(lm.influence(the.linear.model)$hat > 2*p/n)
  lev.top    <- head(rev(sort(lm.influence(the.linear.model)$hat)), topN)
  dffits.out <- which(abs(dffits(the.linear.model)) > 2*sqrt(p/n))
  dffits.top <- head(rev(sort(abs(dffits(the.linear.model)))), topN)
  cooks.out  <- which(cooks.distance(the.linear.model) > 1)
  cooks.top  <- head(rev(sort(cooks.distance(the.linear.model))), topN)
  list(Std.res=res.out, Std.res.top=res.top,
       Leverage=lev.out, Leverage.top=lev.top,
       DFFITS=dffits.out, DFFITS.top=dffits.top,
       Cooks=cooks.out, Cooks.top=cooks.top)
}

out_stats <- show_outliers(m_final, 5)
out_stats
Reduce(intersect, list(out_stats$Std.res, out_stats$DFFITS, out_stats$Leverage))
```

```{r vif}
vif(m_final)
```

# 5. Results Table

```{r results_table}
coefs <- summary(m_final)$coefficients
ci    <- confint(m_final)

results <- data.frame(
  Predictor = rownames(coefs),
  Estimate  = round(coefs[,"Estimate"],   2),
  SE        = round(coefs[,"Std. Error"], 2),
  CI_low    = round(ci[,1], 2),
  CI_high   = round(ci[,2], 2),
  p_value   = round(coefs[,"Pr(>|t|)"],  3)
)

knitr::kable(results,
             col.names = c("Predictor","Estimate","Std. Error",
                           "95% CI (low)","95% CI (high)","p-value"),
             caption = "Final model coefficients",
             row.names = FALSE)

s <- summary(m_final)
cat(sprintf("R-squared: %.3f | Adj R-squared: %.3f | Residual SD: %.2f | N: %d\n",
            s$r.squared, s$adj.r.squared, s$sigma, nrow(df)))
```


```{r}
summary(m_final)
vif(m_final)
```
